# (PART\*) GLMs with proportion data {-}

# Binomial GLMs and proportions

Sometimes, proportion data are more similar to logistic regression than you think.

In discrete counts, we can, for instance, measure the number of presence of individuals in relation to the total number of populations sampled.

We will thus obtain a proportional number of "success" in observing individuals by dividing the counts by the total counts.

In ```glm()```, we have to provide prior weights if the response variable is the proportion of successes.
  
Proportions can be modeled by providing both the number of "successes" and prior weights in the function

```{r, echo = FALSE, eval = TRUE}
mites <- read.csv('mites.csv')
```
```{r, echo = TRUE, eval = TRUE}
prop.reg <- glm(cbind(Galumna, totalabund - Galumna) ~ Topo + WatrCont,
                data = mites,
                family = binomial)
summary(prop.reg)
```

The weights can also be set explicitly in ```glm()```:

```{r, echo = TRUE, eval = TRUE}
prop.reg2 <- glm(prop ~ Topo + WatrCont,
                 data = mites,
                 family = binomial,
                 weights = totalabund)
summary(prop.reg2)
```
